\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-12-02 paper.qmd
Finalizing the scDEED section

**UMAP**

@@

```{r}
result_umap |>
  kableExtra::kable(format = "latex",
                    booktabs = TRUE,
                    longtable = TRUE,
                    label = "scdeedumap") |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-height: 100%
#| fig-cap: "Assessing which of the 2 tSNE layouts with different perplexity ($30$ (red), $320$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents tSNE layout (a, b). Layout b is universally suboptimal. Layout b is the best choice."

>>>>

In the field of single-cell studies, clustering is a common analytical task used to identify groups of cells with similar expression profiles. Non-linear dimensional reduction (NLDR) methods are frequently employed to visualize these clusters and help validate the results. However, it is well known that the 2D embeddings produced by t-SNE and UMAP may not accurately reflect the similarities among cell clusters.

To address this challenge, @xia2023 introduces a statistical method called scDEED, which detects unreliable cell embeddings generated by \gD{} embedding techniques. scDEED calculates a reliability score for each cell embedding based on the similarity between the cell’s \gD{} embedding neighbors and its neighbors prior to embedding. By identifying embeddings with low reliability scores as dubious and those with high scores as trustworthy, scDEED minimizes the number of questionable cell embeddings. This approach also provides clear guidance for optimizing the hyperparameters of an embedding method.

To verify that the \gD{} cell embeddings became better aligned with the cell types after scDEED’s optimization, the Human Peripheral Blood Mononuclear Cells (PBMC) dataset was used. It contained $31,021$ cells with cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

For illustration purposes, we only selected cells generated with inDrops ($n=5858$ cells) and UMAP cell embeddings. Also, @xia2023 used first $50$ principal components to generate the UMAP. The objective is to assess the optimized layout by scDEED, and if it does not accurately represent the three clusters with small separations of the PBMC dataset, then select a reasonable \gD{} layout.

@@

```{r}
#| eval: false
result_umap |>
  kableExtra::kable(format = "latex",
                    booktabs = TRUE,
                    longtable = TRUE,
                    label = "scdeedumap") |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-height: 100%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyperparameter settings (n_neighbors: $30$, min_dist: $0.3$ (red), n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents UMAP layout (a, b). Layout b is universally suboptimal. Layout b is the best choice."

### 2024-12-02 paper.tex
Finalizing the scDEED section

\textbf{UMAP}

\begin{longtable*}{rrr}
\toprule
n\_neighbors & min.dist & number\_dubious\_cells\\
\midrule
5 & 0.0125 & 14\\
6 & 0.0125 & 7\\
7 & 0.0125 & 5\\
8 & 0.0125 & 10\\
9 & 0.0125 & 29\\
\addlinespace
10 & 0.0125 & 7\\
20 & 0.0125 & 9\\
30 & 0.0125 & 10\\
40 & 0.0125 & 4\\
50 & 0.0125 & 3\\
\addlinespace
80 & 0.0125 & 47\\
160 & 0.0125 & 13\\
240 & 0.0125 & 74\\
5 & 0.0500 & 12\\
6 & 0.0500 & 37\\
\addlinespace
7 & 0.0500 & 6\\
8 & 0.0500 & 15\\
9 & 0.0500 & 33\\
10 & 0.0500 & 5\\
20 & 0.0500 & 13\\
\addlinespace
30 & 0.0500 & 4\\
40 & 0.0500 & 4\\
50 & 0.0500 & 8\\
80 & 0.0500 & 15\\
160 & 0.0500 & 55\\
\addlinespace
240 & 0.0500 & 75\\
5 & 0.1000 & 271\\
6 & 0.1000 & 9\\
7 & 0.1000 & 16\\
8 & 0.1000 & 37\\
\addlinespace
9 & 0.1000 & 4\\
10 & 0.1000 & 20\\
20 & 0.1000 & 5\\
30 & 0.1000 & 5\\
40 & 0.1000 & 5\\
\addlinespace
50 & 0.1000 & 25\\
80 & 0.1000 & 4\\
160 & 0.1000 & 4\\
240 & 0.1000 & 13\\
5 & 0.3000 & 45\\
\addlinespace
6 & 0.3000 & 10\\
7 & 0.3000 & 12\\
8 & 0.3000 & 4\\
9 & 0.3000 & 3\\
10 & 0.3000 & 3\\
\addlinespace
20 & 0.3000 & 2\\
30 & 0.3000 & 46\\
40 & 0.3000 & 2\\
50 & 0.3000 & 3\\
80 & 0.3000 & 2\\
\addlinespace
160 & 0.3000 & 2\\
240 & 0.3000 & 21\\
5 & 0.5000 & 88\\
6 & 0.5000 & 3\\
7 & 0.5000 & 5\\
\addlinespace
8 & 0.5000 & 4\\
9 & 0.5000 & 92\\
10 & 0.5000 & 5\\
20 & 0.5000 & 72\\
30 & 0.5000 & 70\\
\addlinespace
40 & 0.5000 & 1\\
50 & 0.5000 & 2\\
80 & 0.5000 & 2\\
160 & 0.5000 & 3\\
240 & 0.5000 & 4\\
\addlinespace
5 & 0.7000 & 13\\
6 & 0.7000 & 86\\
7 & 0.7000 & 95\\
8 & 0.7000 & 29\\
9 & 0.7000 & 95\\
\addlinespace
10 & 0.7000 & 6\\
20 & 0.7000 & 3\\
30 & 0.7000 & 5\\
40 & 0.7000 & 2\\
50 & 0.7000 & 3\\
\addlinespace
80 & 0.7000 & 4\\
160 & 0.7000 & 5\\
240 & 0.7000 & 4\\
5 & 0.9000 & 192\\
6 & 0.9000 & 95\\
\addlinespace
7 & 0.9000 & 233\\
8 & 0.9000 & 228\\
9 & 0.9000 & 101\\
10 & 0.9000 & 7\\
20 & 0.9000 & 3\\
\addlinespace
30 & 0.9000 & 79\\
40 & 0.9000 & 3\\
50 & 0.9000 & 5\\
80 & 0.9000 & 4\\
160 & 0.9000 & 5\\
\addlinespace
240 & 0.9000 & 3\\
\bottomrule
\end{longtable*}

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 tSNE layouts
with different perplexity (\(30\) (red), \(320\) (blue)) on the PBMC
data is the better representation using MSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the left plot and in
the scatterplots represents tSNE layout (a, b). Layout b is universally
suboptimal. Layout b is the best choice.}

>>>>

In the field of single-cell studies, clustering is a common analytical
task used to identify groups of cells with similar expression profiles.
Non-linear dimensional reduction (NLDR) methods are frequently employed
to visualize these clusters and help validate the results. However, it
is well known that the 2D embeddings produced by t-SNE and UMAP may not
accurately reflect the similarities among cell clusters.

To address this challenge, \citet{xia2023} introduces a statistical
method called scDEED, which detects unreliable cell embeddings generated
by \gD{} embedding techniques. scDEED calculates a reliability score for
each cell embedding based on the similarity between the cell's \gD{}
embedding neighbors and its neighbors prior to embedding. By identifying
embeddings with low reliability scores as dubious and those with high
scores as trustworthy, scDEED minimizes the number of questionable cell
embeddings. This approach also provides clear guidance for optimizing
the hyperparameters of an embedding method.

To verify that the \gD{} cell embeddings became better aligned with the
cell types after scDEED's optimization, the Human Peripheral Blood
Mononuclear Cells (PBMC) dataset was used. It contained \(31,021\) cells
with cell type labels, and the gene expression levels were in the unit
of log-transformed UMI count per \(10,000\). They focused on three
sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell
types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

For illustration purposes, we only selected cells generated with inDrops
(\(n=5858\) cells) and UMAP cell embeddings. Also, \citet{xia2023} used
first \(50\) principal components to generate the UMAP. The objective is
to assess the optimized layout by scDEED, and if it does not accurately
represent the three clusters with small separations of the PBMC dataset,
then select a reasonable \gD{} layout.

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 UMAP layouts
with different hyperparameter settings (n\_neighbors: \(30\), min\_dist:
\(0.3\) (red), n\_neighbors: \(80\), min\_dist: \(0.5\) (blue)) on the
PBMC data is the better representation using MSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the left plot and in
the scatterplots represents UMAP layout (a, b). Layout b is universally
suboptimal. Layout b is the best choice.}
